A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-23 03-semantics.tex
Seção 3.3.2, lemas de preservação do \rank

  The auxiliary function~$\pot'$ is defined by
  clauses~\eqref{def.x.pot.first}--\eqref{def.x.pot.last}:
  \begin{enumerate}[(a)]
  \item\label{def.x.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$.
    %%
  \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})
    =\max\{\pot'(p_1),\pot'(p_2)\}$.
    %%
  \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$.
    %%
  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(p_1)+\pot'(p_2)$.
    %%
  \item$\pot'(\ceu{{p_1}\Or{p_2}})=\pot'(p_1)+\pot'(p_2)$.
    %%
  \item If~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
    \begin{align*}
      \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\
      \pot'(\ceu{p_1\AtLoop{p_2}})&=
      \begin{cases}
        \pot'(p_1)              &\text{if\enspace(\dag)}\\
        \pot'(p_1)+\pot'(p_2)   &\text{otherwise}\\
      \end{cases}
    \end{align*}
    where~(\dag) stands for: ``a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ occurs
    in all execution paths of~$p_1$''.
    %%
  \item If~$p_1,p_2\ne\ceu{\Break}$.
    \[
      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
    \]
    %%
  \item\label{def.x.pot.before-last} If~$p_1,p_2\ne\ceu{\Break}$
    and~$p_1,p_2\ne\ceu{\Nop}$,
    \[
      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
    \]
    %%
  \item\label{def.x.pot.last} Otherwise, if none
    of~\eqref{def.pot.first}--\eqref{def.pot.before-last} applies,
    $\pot(\_)=0$.
  \end{enumerate}
\end{definition}

@@

@@

The next two lemmas establish that a single application of~$\out$ or~$\nst$
either preserves or decreases the rank of the input description.  All rank
comparisons assume lexicographical order.

\begin{lemma}\strut
  \label{lem.rank-out}
  \begin{enumerate}[(a)]
  \item\label{lem.rank-out-push} If~$\delta\outpush\delta'$
    then~$\rank(\delta)=\rank(\delta')$.
  \item\label{lem.rank-out-pop} If~$\delta\outpop\delta'$
    then~$\rank(\delta)>\rank(\delta')$.
  \end{enumerate}
\end{lemma}

\begin{lemma}
  \label{lem.rank-nst}
  %%
  If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
\end{lemma}

An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for
transitions~$\trans$ in general.  Such descriptions cannot be advanced
by~$\nst$, as it is nested-irreducible, and neither by~$\outpush$
nor~$\outpop$, as the second coordinate of its rank is~0, which
implies~$e=\nil$ and~$n=0$.

TODO: Lemas e teorema da terminação.  Sketch da prova.  Mencionar que as
restrições sintáticas são fundamentais.

% - there is always a possible transition until n=0
%
% every cannot restart itself
%     - break disallowed
%     - emit ignored

>>>>

  The auxiliary function~$\pot'$ is defined by
  clauses~\eqref{def.x.pot.first}--\eqref{def.x.pot.last}:
  \begin{enumerate}[(a)]
  \item\label{def.x.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$.
    %%
  \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})
    =\max\{\pot'(p_1),\pot'(p_2)\}$.
    %%
  \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$.
    %%
  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(p_1)+\pot'(p_2)$.
    %%
  \item$\pot'(\ceu{{p_1}\Or{p_2}})=\pot'(p_1)+\pot'(p_2)$.
    %%
  \item If~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
    \begin{align*}
      \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\
      \pot'(\ceu{p_1\AtLoop{p_2}})&=
      \begin{cases}
        \pot'(p_1)              &\text{if\enspace(\dag)}\\
        \pot'(p_1)+\pot'(p_2)   &\text{otherwise}\\
      \end{cases}
    \end{align*}
    where~(\dag) stands for: ``a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ occurs
    in all execution paths of~$p_1$''.
    %%
  \item If~$p_1,p_2\ne\ceu{\Break}$.
    \[
      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
    \]
    %%
  \item\label{def.x.pot.before-last} If~$p_1,p_2\ne\ceu{\Break}$
    and~$p_1,p_2\ne\ceu{\Nop}$,
    \[
      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
    \]
    %%
  \item\label{def.x.pot.last} Otherwise, if none
    of~\eqref{def.x.pot.first}--\eqref{def.x.pot.before-last} applies,
    $\pot(\_)=0$.
  \end{enumerate}
\end{definition}

@@

We can now characterize irreducible descriptions in general.

@@

An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for
arbitrary transitions~$\trans$.  Such descriptions cannot be advanced
by~$\nst$, as it is nested-irreducible, and neither by~$\outpush$
nor~$\outpop$, as the second coordinate of its rank is~0, which
implies~$e=\nil$ and~$n=0$.

The next two lemmas establish that a single application of~$\out$ or~$\nst$
either preserves or decreases the rank of the input description.  All rank
comparisons assume lexicographical order, i.e., if~$\rank(\delta)=\<i,j>$
and~$\rank(\delta')=\<i',j'>$ then~$\rank(\delta)>\rank(\delta')$ iff~$i>i'$
or~$i=i$ and~$j>j'$.
%%
The proof of Lemma~\ref{lem.x.rank-out} follows directly from~\R{push}
and~\R{pop} and from Definitions~\ref{def.x.pot} and~\ref{def.x.rank}.  The
proof of Lemma~\ref{lem.x.rank-nst}, however, is by induction on the
structure of~$\nst$ derivations.

\begin{lemma}\strut
  \label{lem.x.rank-out}
  %%
  \begin{enumerate}[(a)]
  \item\label{lem.rank-out-push} If~$\delta\outpush\delta'$
    then~$\rank(\delta)=\rank(\delta')$.
  \item\label{lem.rank-out-pop} If~$\delta\outpop\delta'$
    then~$\rank(\delta)>\rank(\delta')$.
  \end{enumerate}
\end{lemma}

\begin{lemma}
  \label{lem.x.rank-nst}
  %%
  If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
\end{lemma}

### 2018-02-23 06-appendix.tex
Seção 3.3.2, lemas de preservação do \rank

\begin{lemma}
  \label{lem.props-nst-i}\strut\\
  %%
  If~$\<p_1,n,e>\nst[i]\<p_1',n,e'>$ then, for any~$p_2$:
  \begin{enumerate}[(a)]
  \item\label{lem.props-nst-i.a}
    $\<\ceu{p_1;\,p_2},n,e>\nst[i]\<p_1';p_2,n,e'>$;
    %%
  \item\label{lem.props-nst-i.b}
    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[i]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$;
    %%
  \item\label{lem.props-nst-i.c}
    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$;
    %%
  \item\label{lem.props-nst-i.d}
    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}'\AtOr{p_2}},n,e'>$.
  \end{enumerate}
  If~$\<p_2,n,e>\nst[i]\<p_2',n,e'>$, for any~$p_1$ such
  that~$\isblocked(p_1,n)$:
  \begin{enumerate}[(a)]
    \setcounter{enumi}{4}
  \item\label{lem.props-nst-i.e}
    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>$;
    %%
  \item\label{lem.props-nst-i.f}
    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtOr{p_2'}},n,e'>$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  By induction on~$i$.
  %%
  \begin{enumerate}[(a)]
  \item The lemma is trivially true for~$i=0$, as~$p_1=p_1'$, and follows
    directly from~\R{seq-adv} for~$i=1$.  Suppose
    \begin{equation}
      \label{lem.props-nst-i.a.eq1}
      \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[i-1]\<p_1',n,e'>\,,
    \end{equation}
    for some~$i>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible, i.e.,
    $e=\nil$ and~$p\ne{\ceu{\Nop},\ceu{\Break}}$ and~$\isblocked(p_1'',n)$
    is false.  By~\eqref{lem.props-nst-i.a.eq1} and by~\R{seq-adv},
    \begin{equation}
      \label{lem.props-nst-i.a.eq2}
      \<\ceu{p_1;\,p_2},n,e>\nst[1]\<\ceu{p_1'';\,p_2},n,e''>\,.
    \end{equation}
    From~\eqref{lem.props-nst-i.a.eq1}, by the induction hypothesis,
    \begin{equation}
      \label{lem.props-nst-i.a.eq3}
      \<\ceu{p_1'';\,p_2},n,e''>\nst[i-1]\<\ceu{p_1';\,p_2},n,e'>\,.
    \end{equation}
    From~\eqref{lem.props-nst-i.a.eq2} and~\eqref{lem.props-nst-i.a.eq3},
    \[
      \<\ceu{p_1;\,p_2},n,e>\nst[i]\<\ceu{p_1';\,p_2},n,e'>\,.
    \]

@@

The next two lemmas establish that a single application of~$\out$ or~$\nst$
either preserves or decreases the rank of the input description.  All rank
comparisons assume lexicographical order.

\begin{lemma}\strut
  \label{lem.rank-out}
  \begin{enumerate}[(a)]
  \item\label{lem.rank-out-push} If~$\delta\outpush\delta'$
    then~$\rank(\delta)=\rank(\delta')$.
  \item\label{lem.rank-out-pop} If~$\delta\outpop\delta'$
    then~$\rank(\delta)>\rank(\delta')$.
  \end{enumerate}
\end{lemma}
\begin{proof}\strut
  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$,
  $\rank(\delta)=\<i,j>$, and~$\rank(\delta')=\<i',j'>$.
  \begin{enumerate}[(a)]
  \item Suppose~$\<p,n,e>\outpush\<p',n',e'>$.  Then, by~\R{push},
    $e\ne\nil$, $p'=\bcast(p,e)$, $n'=n+1$, and~$e'=\nil$.  By
    Definition~\ref{def.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
    as~$e'=\nil$ and~$n'=n+1$; hence~$j=j'$.
    %%
    It remains to be shown that~$i=i'$:
    \begin{align*}
      i&=\pot(p,e)
         \tag*{by Definition~\ref{def.rank}}\\
       &=\pot'(\bcast(p,e))
         \tag*{by Definition~\ref{def.pot}}\\
       &=\pot'(p')
         \tag*{since~$p'=\bcast(p,e)$}\\
       &=\pot'(\bcast(p',\nil))
         \tag*{by definition of~$\bcast$}\\
       &=\pot'(\bcast(p',e'))
         \tag*{since~$e'=\nil$}\\
       &=\pot(p',e')
         \tag*{by Definition~\ref{def.pot}}\\
       &=i'
         \tag*{by Definition~\ref{def.rank}}
    \end{align*}
    Therefore, $\<i,j>=\<i',j'>$.

@@

An irreducible description~$\delta_\#=\<p,n,e>$ serves as a normal form for
general transitions~$\trans$.  Such descriptions cannot be advanced
by~$\nst$, as it is nested-irreducible, and neither by~$\outpush$
nor~$\outpop$, as the second coordinate of its rank is~0, which
implies~$e=\nil$ and~$n=0$.

>>>>

\begin{lemma}
  \label{lem.props-nst-i}\strut\\
  %%
  If~$\<p_1,n,e>\nst[i]\<p_1',n,e'>$ then, for any~$p_2$:
  \begin{enumerate}[(a)]
  \item\label{lem.props-nst-i.a}
    $\<\ceu{p_1;\,p_2},n,e>\nst[i]\<p_1';p_2,n,e'>$.
    %%
  \item\label{lem.props-nst-i.b}
    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[i]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$.
    %%
  \item\label{lem.props-nst-i.c}
    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$.
    %%
  \item\label{lem.props-nst-i.d}
    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}'\AtOr{p_2}},n,e'>$.
  \end{enumerate}
  If~$\<p_2,n,e>\nst[i]\<p_2',n,e'>$, for any~$p_1$ such
  that~$\isblocked(p_1,n)$:
  \begin{enumerate}[(a)]
    \setcounter{enumi}{4}
  \item\label{lem.props-nst-i.e}
    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>$.
    %%
  \item\label{lem.props-nst-i.f}
    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtOr{p_2'}},n,e'>$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  By induction on~$i$.
  %%
  \begin{enumerate}[(a)]
  \item The lemma is trivially true for~$i=0$, as~$p_1=p_1'$, and follows
    directly from~\R{seq-adv} for~$i=1$.  Suppose
    \begin{equation}
      \label{lem.props-nst-i.a.eq1}
      \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[i-1]\<p_1',n,e'>\,,
    \end{equation}
    for some~$i>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible, i.e.,
    $e=\nil$ and~$p\ne{\ceu{\Nop},\ceu{\Break}}$ and~$\isblocked(p_1'',n)$
    is false.  By~\eqref{lem.props-nst-i.a.eq1} and by~\R{seq-adv},
    \begin{equation}
      \label{lem.props-nst-i.a.eq2}
      \<\ceu{p_1;\,p_2},n,e>\nst[1]\<\ceu{p_1'';\,p_2},n,e''>\,.
    \end{equation}
    From~\eqref{lem.props-nst-i.a.eq1}, by the induction hypothesis,
    \begin{equation}
      \label{lem.props-nst-i.a.eq3}
      \<\ceu{p_1'';\,p_2},n,e''>\nst[i-1]\<\ceu{p_1';\,p_2},n,e'>\,.
    \end{equation}
    From~\eqref{lem.props-nst-i.a.eq2} and~\eqref{lem.props-nst-i.a.eq3},
    \[
      \<\ceu{p_1;\,p_2},n,e>\nst[i]\<\ceu{p_1';\,p_2},n,e'>\,.
    \]

@@

\begin{lemma}\strut
  \label{lem.rank-out}
  %%
  \begin{enumerate}[(a)]
  \item\label{lem.rank-out-push} If~$\delta\outpush\delta'$
    then~$\rank(\delta)=\rank(\delta')$.
  \item\label{lem.rank-out-pop} If~$\delta\outpop\delta'$
    then~$\rank(\delta)>\rank(\delta')$.
  \end{enumerate}
\end{lemma}
\begin{proof}\strut
  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$,
  $\rank(\delta)=\<i,j>$, and~$\rank(\delta')=\<i',j'>$.
  \begin{enumerate}[(a)]
  \item Suppose~$\<p,n,e>\outpush\<p',n',e'>$.  Then, by~\R{push},
    $e\ne\nil$, $p'=\bcast(p,e)$, $n'=n+1$, and~$e'=\nil$.  By
    Definition~\ref{def.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
    as~$e'=\nil$ and~$n'=n+1$; hence~$j=j'$.
    %%
    It remains to be shown that~$i=i'$:
    \begin{align*}
      i&=\pot(p,e)
         \tag*{by Definition~\ref{def.rank}}\\
       &=\pot'(\bcast(p,e))
         \tag*{by Definition~\ref{def.pot}}\\
       &=\pot'(p')
         \tag*{since~$p'=\bcast(p,e)$}\\
       &=\pot'(\bcast(p',\nil))
         \tag*{by definition of~$\bcast$}\\
       &=\pot'(\bcast(p',e'))
         \tag*{since~$e'=\nil$}\\
       &=\pot(p',e')
         \tag*{by Definition~\ref{def.pot}}\\
       &=i'
         \tag*{by Definition~\ref{def.rank}}
    \end{align*}
    Therefore, $\<i,j>=\<i',j'>$.

@@
